Finding a closest point in a lattice of Voronoi's first kind

### 2014-01-02 paper.tex
pulled out arguments from Lemmas 4.10 and 4.11 (now 4.12 and 4.13)

An $n$-dimensional \term{lattice} $\Lambda$ is a discrete set of vectors from $\reals^m$, $m \geq n$, formed by the integer linear combinations of a set of linearly independent basis vectors $b_1, \dots, b_n$ from $\reals^m$~\cite{SPLAG}.  That is, $\Lambda$ consists of all those vectors, or \emph{lattice points}, $x \in \reals^m$ satisfying
\[
  x = b_1 u_1 + b_2u_2 + \dots + b_n u_n \qquad u_1, \dots , u_n \in \ints. 
\] 
%In this paper vectors are assumed to be column vectors unless otherwise stated.
Given a lattice $\Lambda$ in $\reals^m$ and a vector $y \in \reals^m$, a problem of interest is to find a lattice point $x \in \Lambda$ such that the squared Euclidean norm
\[
\| y - x \|^2 = \sum_{i=1}^m (y_i - x_i)^2
\] 
is minimised.  This is called the \emph{closest lattice point problem} (or \emph{closest vector problem}) and a solution is called a \emph{closest lattice point} (or simply \emph{closest point}) to $y$. %Equivalently we wish to minimise
%\[
%\| y - \sum_{i=1}^n b_i w_i \|^2
%\]
%over integers $w_1,\dots,w_n$.  
A related problem is to find a lattice point of minimum nonzero Euclidean length, that is, a lattice point of length
\[
\min_{x\in \Lambda \backslash \{ \zerobf \} } \| x \|^2,
\]
where $\Lambda \backslash  \{\zerobf\}$ denotes the set of lattice point not equal to the origin $\zerobf$.  This is called the \emph{shortest vector problem}.

@@

\begin{proposition}\label{eq:latticepointsinsphere}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice with packing radius $\rho$.  Let $S$ be an $m$-dimensional hypersphere of radius $r$ centered at $t \in \reals^{m}$.  The number of lattice points from $\Lambda$ in the sphere $S$ is less than $\ceil{r/\rho}^{n}$ where $\ceil{\cdot}$ denotes the smallest integer strictly larger than its argument. 
 \end{proposition}
 \begin{proof}
Since the packing radius $\rho$ is the point on the boundary of the Voronoi that is closest to the origin, the hypersphere $S$ is a subset of the Voronoi cell scaled by $r/\rho$ and translated by $t$.  That is, $S \subset r/\rho \vor(\Lambda) + t$.  The proof follows because the number of lattice points in $r/\rho \vor(\Lambda) + t$ is less than $\ceil{r/\rho}^n$ (this follows from, for example, \cite[Theorem~2]{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases}, see also~\cite[Lemma~3.7]{Micciancio09adeterministic}).
\end{proof}

@@

Two factors contribute to the computational complexity of this iterative approach to compute a closest lattice point.  The first factor is computing the minimum over the set $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer}.  In general a lattice can have as many as $2^{n+1}-2$ relevant vectors and so computing a minimiser directly can require a number of operations that grows exponentially with $n$.  To add to this it is typically the case that the set of relevant vectors $\relevant(\Lambda)$ must be stored in memory, and so the algorithm can require an amount of memory that grows exponentially with $n$.  We will show that, for a lattice of Voronoi's first kind, the set of relevant vectors has a compact representation in terms of what is called its \emph{obtuse superbasis}.  To store the obtuse superbasis requires an amount of memory of order $O(n^2)$ in the worst case.  We also show that for a lattice of Voronoi's first kind the minimisation over $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer} can be solved efficiently by computing a minimum cut in an undirected flow network.  Using known algorithms a minimiser can be computed in $O(n^3)$ operations~\cite{Goldberg:1986:NAM:12130.12144,EdmondsKarp_max_flow,Cormen2001}.

@@

We are interested in solving the closest lattice point problem for lattices of Voronoi's first kind.  Let $\Lambda \subset \reals^m$ be an $n$ dimensional lattice of Voronoi's first kind with obtuse superbasis $b_1,\dots,b_{n+1}$ and let $y \in \reals^m$.  We want to find $n$ integers $w_1,\dots,w_n$ that minimise
\[
\| y - \sum_{i=1}^n b_i w_i \|^2.
\]
We can equivalently find $n+1$ integers $w_1,\dots,w_{n+1}$ that minimise
\[
\| y - \sum_{i=1}^{n+1} b_i w_i \|^2.
\]
We will use the iterative procedure described in~\eqref{eq:relvectsminimslicer} to do this.  In what follows we will assume that $y$ lies in the space spanned by the basis vectors $b_1,\dots,b_{n}$.  This assumption is without loss of generality because $x$ is closest lattice point to $y$ if and only if $x$ is a the closest lattice point to the projection of $y$ orthogonal into the space spanned by $b_1,\dots,b_{n+1}$.  Let
\begin{equation}\label{eq:matrxBobtusebasis}
B = (b_1\,\,b_2\,\,\dots\,\,b_{n+1})
\end{equation}
be the matrix with columns given by $b_1,\dots,b_{n+1}$ and let $z \in \reals^{n+1}$ be a column vector such that $y = Bz$.  We now want to find a column vector $w = (w_1,\dots,w_{n+1})^\prime$ of integers such that
\begin{equation}\label{eq:tominimise}
\| B(z  -  w) \|^2
\end{equation}
is minimised.  Define the column vector $u_0 = \floor{z}$ where $\floor{\cdot}$ denotes the largest integer less than or equal to its argument and operates on vectors elementwise. In view of Theorem~\ref{thm:revvecssuperbase} the iterative procedure~\eqref{eq:relvectsminimslicer} to compute a closest lattice point can be written in the form
\begin{align}
x_{k+1} &= B u_{k+1} \label{eq:xseqfirsttype}  \\
u_{k+1} &= u_k + p_k \nonumber \\
p_k &= \arg\min_{p \in \{0,1\}^{n+1}}\| B(z - u_k - p) \|^2, \label{eq:pvecmin}
\end{align}
where $\{0,1\}^{n+1}$ denotes the set of column vectors of length $n+1$ with elements equal to zero or one.  Observe that the procedure is initialised at the lattice point $x_0 = Bu_0 = B\floor{z}$.  This choice of initial lattice point is important.  In Section~\ref{sec:comp-clos-relev} we show how the minimisation over $\{0,1\}^{n+1}$ in~\eqref{eq:pvecmin} can be computed efficiently in $O(n^3)$ operations by computing a minimum cut in an undirected flow network.  In the remainder of this section we prove that this iterative procedure results in a closest lattice point after at most $n$ iterations.  That is, we show that there exists a positive integer $K \leq n$ such that $x_K$ is a closest lattice point to $y = Bz$.

@@

It is necessary to introduce some notation.  Given a subset $S$ of indices from $\{1,\dots,n+1\}$ we denote by $I_S$ the vector of length $n+1$ with elements equal to zero except for those with indices from $S$ that are equal to one.  That is, if $g = I_S$ then $g$ is the vector with elements
\[
g_i = \begin{cases}
1 & i \in S\\
0 & i \notin S.
\end{cases}
\]
If $p \in \reals^{n+1}$ then $p - I_S$ denotes the vector $p$ with those elements with index from $S$ decremented by one.  Similarly, $p + I_S$ denotes the vector $p$ with those elements with index from $S$ incremented by one.

\begin{lemma}\label{lem:decSellings}
Let $p \in \reals^{n+1}$, let $S$ be a subset of the indices of $p$, and let $B$ be the matrix from~\eqref{eq:matrxBobtusebasis} with column given by the superbasis vectors $b_1,\dots,b_{n+1}$.  The following equalities hold:
\begin{enumerate}
\item  ${\displaystyle \|Bp\|^2 - \|B(p - I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i)}$, \label{eq:lem:decSellingsdec}
\item  ${\displaystyle \|Bp\|^2 - \|B(p + I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j)}$, \label{eq:lem:decSellingsinc}
\end{enumerate}
where $q_{ij} = \dotprod{b_i}{b_j}$ are the Selling parameters.
\end{lemma}
\begin{proof}
We give a proof for part~\ref{eq:lem:decSellingsdec}.  The proof for part~\ref{eq:lem:decSellingsinc} is similar.  Observe that
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_i^2 =  \sum_{i=1}^{n+1}p_i^2\sum_{j=1}^{n+1} q_{ij} = 0
\]
and
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_j^2 =  \sum_{j=1}^{n+1}p_j^2\sum_{i=1}^{n+1} q_{ij} = 0
\]
since $\sum_{i=1}^{n+1} q_{ij} = 0$ as a result of the superbasis condition~\eqref{eq:superbasecond}.  Now
\begin{align*}
\|Bp\|^2 &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} p_i p_j \\
&= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (\tfrac{1}{2}p_j^2 + p_i p_j -  \tfrac{1}{2}p_i^2) \\
&= -\frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (p_i - p_j)^2.
\end{align*}
This is closely related to what is known as Sellings formula~\cite[Proposition 2.3.1]{Valentin2003_coverings_tilings_low_dimension}~\cite{Selling1874}.  Put $g = p - I_S$.  It follows that
\[
\|Bp\|^2 - \|Bg\|^2 = \frac{1}{2} \sum_{i=1}^{n+1}\sum_{j=1}^{n+1}q_{ij}d_{ij},
\]
where $d_{ij} = (g_i - g_j)^2 - (p_i - p_j)^2$.  If both $i \notin S$ and $j \notin S$ then $p_i = g_i$, $p_j = g_j$ and so $d_{ij} = 0$.  Similarly if both $i \in S$ and $j \in S$ then 
\[
d_{ij} = (p_i-1 - p_j+1)^2 - (p_i - p_j)^2 = 0.
\]
If $i \in S$ and $j \notin S$ then
\[
d_{ij} = (p_i-1 - p_j)^2 - (p_i - p_j)^2= 1 + 2p_j - 2p_i,
\]
while if $i \notin S$ and $j \in S$ then
\[
d_{ij} = (p_i - p_j+1)^2 - (p_i - p_j)^2 = 1 + 2p_i - 2p_j.
\]
We thus have
\begin{align*}
\|Bp\|^2 - \|Bg\|^2 &= \frac{1}{2} \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \frac{1}{2} \sum_{i \notin S}\sum_{j \in S}q_{ij}(1 + 2p_i - 2p_j) \\
&= \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i),
\end{align*}
where the last line holds because $q_{ij} = q_{ji}$.
\end{proof}

@@

\begin{lemma}\label{lem:rngdecreases}
Let $Bu$ with $u \in \ints^{n+1}$ be a lattice point that is $\ell$-close to $y = Bz$ where $\ell > 0$.  Let $g \in \{0,1\}^{n+1}$ be such that
\begin{equation}\label{eq:qismin}
\|B(z - u - g)\|^2 = \min_{p \in \{0,1\}^{n+1}}\|B(z - u - p)\|^2.
\end{equation}
The lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
\end{lemma}
\begin{proof}
Because $Bu$ is $\ell$-close to $y$ there exists $v \in \ints^{n+1}$ with $\rng(v) = \ell$ such that $B(u+v)$ is a closest lattice point to $y$.  Let $A$ be the (possibly empty) subset of indices $\{1,\dots,n+1\}$ such that $g_i = 1$ and $v_i = \min(v)$ and let $D$ be the (possibly empty) subset of indices such that $g_i = 0$ and $v_i = \max(v)$. That is,
\[
A = \{i \mid g_i = 1, v_i = \min(v) \}, \qquad D = \{i \mid g_i = 0, v_i = \max(v) \}.
\]
Put $d = g - I_A + I_D$ and put $w = v - d$.  Observe that $d_i = 0$ whenever $v_i = \min(v)$ and so $\min(w) = \min(v-d) = \min(v)$.  Also, $d_i = 1$ whenever $v_i = \max(v)$ and so $\max(w) = \max(v-d) = \max(v) - 1$.  Thus,
\[
\rng(w) = \max(v) - 1 - \min(v) = \rng(v) - 1.
\]
The lemma will follow if we show that $B(u+g+w)$ is a closest lattice point to $y$ since then $B(u+g)$ with be $(\ell-1)$-close to $y$.  We show this in two stages.  First put $h = v - g + I_A$.  Suppose that $B(u+g+h)$ is not a closest lattice point to $y = Bz$.  That is, suppose that 
\[
\|B(z-u-g-h)\|^2 = \|B(z-u-v-I_A)\|^2 > \|B(z-u-v)\|^2.
\]
In this case, Lemma~\ref{lem:vqcontramin} (proved below) with $t = z-u$ and $S = A$ shows that 
\begin{equation}\label{eq:lem:rng:contag}
\|B(z-u - g)\|^2 > \min_{p \in \{0,1\}^{n+1}}\|B(z- u - p)\|^2,
\end{equation}
contradicting~\eqref{eq:qismin}.  Thus, $B(u+g+h)$ is a closest lattice point to $y$.  Now suppose that $B(u+g+w) = B(u+g+h-I_D)$ is not a closest lattice point to $y$.  That is, suppose that
\[
\|B(z-u-g-h+I_D)\|^2 > \|B(z-u-h-g)\|^2.
\]
Lemma~\ref{lem:vqcontramax} (proved below) with $t = z-u$ and $S = D$ shows that~\eqref{eq:lem:rng:contag} occurs again, contradicting~\eqref{eq:qismin}.  Thus, $B(u+g+w)$ is a closest lattice point to $y$ and, since $\rng(w) = \ell-1$, the lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
 \end{proof}

@@

\begin{theorem}
Let $x_0,x_1,\dots$ be the sequence of lattice points given by the iterative procedure~\eqref{eq:xseqfirsttype}.  There exists $K \leq n$ such that $x_K$ is a closest lattice point to $y = Bz$.
\end{theorem}
\begin{proof}
Let $x_k = B u_k$ with $u_k \in \ints^{n+1}$ be the lattice point obtained on the $k$th iteration of the procedure.  Suppose that $x_k$ is $\ell$-close to $y=Bz$ with $\ell > 0$.  The procedure computes $p_{k} \in \{0,1\}^{n+1}$ satisfying
\[
\|B(z - u_k - p_{k})\|^2 = \min_{p \in \{0,1\}^{n+1}}\|B(z - u_k - p)\|^2
\]
and puts $x_{k+1} = B(u_k + p_k)$.  It follows from Lemma~\ref{lem:rngdecreases} that $x_{k+1}$ is $(\ell-1)$-close to $y$.  By Lemma~\ref{lem:roundzclose} the lattice point that initialises the procedure $x_0 = B \floor{z}$ is $K$-close to $y$ where $K \leq n$.  Thus, $x_1$ is $(K-1)$-close, $x_2$ is $(K-2)$-close, and so on until $x_K$ is $0$-close.  That $x_K$ is a closest lattice point to $y$ follows from Lemma~\ref{lem:rngzeroclosestpoint}.
\end{proof}

We now provide proof of Lemmas~\ref{lem:vqcontramin} and~\ref{lem:vqcontramax} that were used in the proof of Lemma~\ref{lem:rngdecreases}.

>>>>

An $n$-dimensional \term{lattice} $\Lambda$ is a discrete set of vectors from $\reals^m$, $m \geq n$, formed by the integer linear combinations of a set of linearly independent basis vectors $b_1, \dots, b_n$ from $\reals^m$~\cite{SPLAG}.  That is, $\Lambda$ consists of all those vectors, or \emph{lattice points}, $x \in \reals^m$ satisfying
\[
  x = b_1 u_1 + b_2u_2 + \dots + b_n u_n \qquad u_1, \dots , u_n \in \ints. 
\] 
%In this paper vectors are assumed to be column vectors unless otherwise stated.
Given a lattice $\Lambda$ in $\reals^m$ and a vector $y \in \reals^m$, a problem of interest is to find a lattice point $x \in \Lambda$ such that the squared Euclidean norm
\[
\| y - x \|^2 = \sum_{i=1}^m (y_i - x_i)^2
\] 
is minimised.  This is called the \emph{closest lattice point problem} (or \emph{closest vector problem}) and a solution is called a \emph{closest lattice point} (or simply \emph{closest point}) to $y$. %Equivalently we wish to minimise
%\[
%\| y - \sum_{i=1}^n b_i w_i \|^2
%\]
%over integers $w_1,\dots,w_n$.  
A related problem is to find a lattice point of minimum nonzero Euclidean length, that is, a lattice point of length
\[
\min_{x\in \Lambda \backslash \{ \zerobf \} } \| x \|^2,
\]
where $\Lambda \backslash  \{\zerobf\}$ denotes the set of lattice points not equal to the origin $\zerobf$.  This is called the \emph{shortest vector problem}.

@@

\begin{proposition}\label{eq:latticepointsinsphere}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice with packing radius $\rho$.  Let $S$ be an $m$-dimensional hypersphere of radius $r$ centered at $t \in \reals^{m}$.  The number of lattice points from $\Lambda$ in the sphere $S$ is less than $\ceil{r/\rho}^{n}$ where $\ceil{\cdot}$ denotes the smallest integer strictly larger than its argument. 
 \end{proposition}
 \begin{proof}
The packing radius $\rho$ is the point on the boundary of the Voronoi cell $\vor(\Lambda)$ that is closest to the origin. Therefore, the sphere $S$ is a subset of $\vor(\Lambda)$ scaled by $r/\rho$ and translated by $t$.  That is, $S \subset r/\rho \vor(\Lambda) + t$.  The proof follows because the number of lattice points in $r/\rho \vor(\Lambda) + t$ is less than $\ceil{r/\rho}^n$ (this follows from, for example, \cite[Theorem~2]{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases}, see also~\cite[Lemma~3.7]{Micciancio09adeterministic}).
\end{proof}

@@

Two factors contribute to the computational complexity of this iterative approach to compute a closest lattice point.  The first factor is computing the minimum over the set $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer}.  In general a lattice can have as many as $2^{n+1}-2$ relevant vectors and so computing a minimiser directly can require a number of operations that grows exponentially with $n$.  To add to this it is typically the case that the set of relevant vectors $\relevant(\Lambda)$ must be stored in memory, and so the algorithm can require an amount of memory that grows exponentially with $n$.  We will show that for a lattice of Voronoi's first kind the set of relevant vectors has a compact representation in terms of what is called its \emph{obtuse superbasis}.  To store the obtuse superbasis requires an amount of memory of order $O(n^2)$ in the worst case.  We also show that for a lattice of Voronoi's first kind the minimisation over $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer} can be solved efficiently by computing a minimum cut in an undirected flow network.  Using known algorithms a minimiser can be computed in $O(n^3)$ operations~\cite{Goldberg:1986:NAM:12130.12144,EdmondsKarp_max_flow,Cormen2001}.

@@

We are interested in solving the closest lattice point problem for lattices of Voronoi's first kind.  Let $\Lambda \subset \reals^m$ be an $n$ dimensional lattice of Voronoi's first kind with obtuse superbasis $b_1,\dots,b_{n+1}$ and let $y \in \reals^m$.  We want to find $n$ integers $w_1,\dots,w_n$ that minimise
\[
\| y - \sum_{i=1}^n b_i w_i \|^2.
\]
We can equivalently find $n+1$ integers $w_1,\dots,w_{n+1}$ that minimise
\[
\| y - \sum_{i=1}^{n+1} b_i w_i \|^2.
\]
We will use the iterative procedure described in~\eqref{eq:relvectsminimslicer} to do this.  In what follows we will assume that $y$ lies in the space spanned by the basis vectors $b_1,\dots,b_{n}$.  This assumption is without loss of generality because $x$ is closest lattice point to $y$ if and only if $x$ is a the closest lattice point to the projection of $y$ orthogonal into the space spanned by $b_1,\dots,b_{n}$.  Let
\begin{equation}\label{eq:matrxBobtusebasis}
B = (b_1\,\,b_2\,\,\dots\,\,b_{n+1})
\end{equation}
be the matrix with columns given by $b_1,\dots,b_{n+1}$ and let $z \in \reals^{n+1}$ be a column vector such that $y = Bz$.  We now want to find a column vector $w = (w_1,\dots,w_{n+1})^\prime$ of integers such that
\begin{equation}\label{eq:tominimise}
\| B(z  -  w) \|^2
\end{equation}
is minimised.  Define the column vector $u_0 = \floor{z}$ where $\floor{\cdot}$ denotes the largest integer less than or equal to its argument and operates on vectors elementwise. In view of Theorem~\ref{thm:revvecssuperbase} the iterative procedure~\eqref{eq:relvectsminimslicer} to compute a closest lattice point can be written in the form
\begin{align}
x_{k+1} &= B u_{k+1} \label{eq:xseqfirsttype}  \\
u_{k+1} &= u_k + p_k \nonumber \\
p_k &= \arg\min_{p \in \{0,1\}^{n+1}}\| B(z - u_k - p) \|^2, \label{eq:pvecmin}
\end{align}
where $\{0,1\}^{n+1}$ denotes the set of column vectors of length $n+1$ with elements equal to zero or one.  Observe that the procedure is initialised at the lattice point $x_0 = Bu_0 = B\floor{z}$.  This choice of initial lattice point is important.  In Section~\ref{sec:comp-clos-relev} we show how the minimisation over $\{0,1\}^{n+1}$ in~\eqref{eq:pvecmin} can be computed efficiently in $O(n^3)$ operations by computing a minimum cut in an undirected flow network.  In the remainder of this section we prove that this iterative procedure results in a closest lattice point after at most $n$ iterations.  That is, we show that there exists a positive integer $K \leq n$ such that $x_K$ is a closest lattice point to $y = Bz$.

@@

It is necessary to introduce some notation.  For $S$ and subset of the indices $\{1,\dots,n+1\}$ let $I_S$ denote the column vector of length $n+1$ with $i$th element equal to one if $i \in S$ and $0$ otherwise.

\begin{lemma}\label{lem:decSellings}
Let $p \in \reals^{n+1}$, let $S$ be a subset of the indices of $p$, and let $B$ be the matrix from~\eqref{eq:matrxBobtusebasis} with columns given by the superbasis vectors $b_1,\dots,b_{n+1}$.  The following equalities hold:
\begin{enumerate}
\item  ${\displaystyle \|Bp\|^2 - \|B(p - I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i)}$, \label{eq:lem:decSellingsdec}
\item  ${\displaystyle \|Bp\|^2 - \|B(p + I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j)}$, \label{eq:lem:decSellingsinc}
\end{enumerate}
where $q_{ij} = \dotprod{b_i}{b_j} \leq 0$ are the Selling parameters.
\end{lemma}
\begin{proof}
We give a proof for part~\ref{eq:lem:decSellingsdec}.  The proof for part~\ref{eq:lem:decSellingsinc} is similar.  Observe that
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_i^2 =  \sum_{i=1}^{n+1}p_i^2\sum_{j=1}^{n+1} q_{ij} = 0
\]
and
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_j^2 =  \sum_{j=1}^{n+1}p_j^2\sum_{i=1}^{n+1} q_{ij} = 0
\]
since $\sum_{i=1}^{n+1} q_{ij} = \sum_{j=1}^{n+1} q_{ij} = 0$ as a result of the superbasis condition~\eqref{eq:superbasecond}.  Now
\begin{align*}
\|Bp\|^2 &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} p_i p_j \\
&= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (\tfrac{1}{2}p_j^2 + p_i p_j -  \tfrac{1}{2}p_i^2) \\
&= -\frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (p_i - p_j)^2.
\end{align*}
This is closely related to what is known as Sellings formula~\cite[Proposition 2.3.1]{Valentin2003_coverings_tilings_low_dimension}~\cite{Selling1874}.  Put $g = p - I_S$.  It follows that
\[
\|Bp\|^2 - \|Bg\|^2 = \frac{1}{2} \sum_{i=1}^{n+1}\sum_{j=1}^{n+1}q_{ij}d_{ij},
\]
where $d_{ij} = (g_i - g_j)^2 - (p_i - p_j)^2$.  If both $i \notin S$ and $j \notin S$ then $p_i = g_i$, $p_j = g_j$ and so $d_{ij} = 0$.  Similarly if both $i \in S$ and $j \in S$ then 
\[
d_{ij} = (p_i-1 - p_j+1)^2 - (p_i - p_j)^2 = 0.
\]
If $i \in S$ and $j \notin S$ then
\[
d_{ij} = (p_i-1 - p_j)^2 - (p_i - p_j)^2= 1 + 2p_j - 2p_i,
\]
while if $i \notin S$ and $j \in S$ then
\[
d_{ij} = (p_i - p_j+1)^2 - (p_i - p_j)^2 = 1 + 2p_i - 2p_j.
\]
We thus have
\begin{align*}
\|Bp\|^2 - \|Bg\|^2 &= \frac{1}{2} \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \frac{1}{2} \sum_{i \notin S}\sum_{j \in S}q_{ij}(1 + 2p_i - 2p_j) \\
&= \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i),
\end{align*}
where the last line holds because $q_{ij} = q_{ji}$.
\end{proof}

@@

\begin{lemma}\label{lem:rngdecreases}
Let $Bu$ with $u \in \ints^{n+1}$ be a lattice point that is $\ell$-close to $y = Bz$ where $\ell > 0$.  Let $g \in \{0,1\}^{n+1}$ be such that
\begin{equation}\label{eq:qismin}
\|B(z - u - g)\|^2 = \min_{p \in \{0,1\}^{n+1}}\|B(z - u - p)\|^2.
\end{equation}
The lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
\end{lemma}
\begin{proof}
Because $Bu$ is $\ell$-close to $y$ there exists $v \in \ints^{n+1}$ with $\rng(v) = \ell$ such that $B(u+v)$ is a closest lattice point to $y$.  Let $A$ be the (possibly empty) subset of indices $\{1,\dots,n+1\}$ such that $g_i = 1$ and $v_i = \min(v)$ and let $D$ be the (possibly empty) subset of indices such that $g_i = 0$ and $v_i = \max(v)$. That is,
\[
A = \{i \mid g_i = 1, v_i = \min(v) \}, \qquad D = \{i \mid g_i = 0, v_i = \max(v) \}.
\]
Put $d = g - I_A + I_D$ and put $w = v - d$.  Observe that $d_i = 0$ whenever $v_i = \min(v)$ and so $\min(w) = \min(v-d) = \min(v)$.  Also, $d_i = 1$ whenever $v_i = \max(v)$ and so $\max(w) = \max(v-d) = \max(v) - 1$.  Thus,
\[
\rng(w) = \max(v) - 1 - \min(v) = \rng(v) - 1 = \ell - 1.
\]
The lemma will follow if we show that $B(u+g+w)$ is a closest lattice point to $y$ since then the lattice point $B(u+g)$ with be $(\ell-1)$-close to $y$.  We show this in two stages.  First put $h = v - g + I_A$.  Suppose that $B(u+g+h)$ is not a closest lattice point to $y = Bz$.  That is, suppose that 
\[
\|B(z-u-g-h)\|^2 = \|B(z-u-v-I_A)\|^2 > \|B(z-u-v)\|^2.
\]
In this case, Lemma~\ref{lem:vqcontramin} (proved below) with $t = z-u$ and $S = A$ shows that 
\begin{equation}\label{eq:lem:rng:contag}
\|B(z-u - g)\|^2 > \min_{p \in \{0,1\}^{n+1}}\|B(z- u - p)\|^2,
\end{equation}
contradicting~\eqref{eq:qismin}.  Thus, $B(u+g+h)$ is a closest lattice point to $y$.  Now suppose that $B(u+g+w) = B(u+g+h-I_D)$ is not a closest lattice point to $y$.  That is, suppose that
\[
\|B(z-u-g-h+I_D)\|^2 > \|B(z-u-g-h)\|^2.
\]
Lemma~\ref{lem:vqcontramax} (proved below) with $t = z-u$ and $S = D$ shows that~\eqref{eq:lem:rng:contag} occurs again, contradicting~\eqref{eq:qismin}.  Thus, $B(u+g+w)$ is a closest lattice point to $y$ and, since $\rng(w) = \ell-1$, the lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
 \end{proof}

@@

\begin{theorem}
Let $x_0,x_1,\dots$ be the sequence of lattice points given by the iterative procedure~\eqref{eq:xseqfirsttype}.  There exists $K \leq n$ such that $x_K$ is a closest lattice point to $y = Bz$.
\end{theorem}
\begin{proof}
Let $x_k = B u_k$ be the lattice point obtained on the $k$th iteration of the procedure.  Suppose that $x_k$ is $\ell$-close to $y=Bz$ with $\ell > 0$.  The procedure computes $p_{k} \in \{0,1\}^{n+1}$ satisfying
\[
\|B(z - u_k - p_{k})\|^2 = \min_{p \in \{0,1\}^{n+1}}\|B(z - u_k - p)\|^2
\]
and puts $x_{k+1} = B(u_k + p_k)$.  It follows from Lemma~\ref{lem:rngdecreases} that $x_{k+1}$ is $(\ell-1)$-close to $y$.  By Lemma~\ref{lem:roundzclose} the lattice point that initialises the procedure $x_0 = B \floor{z}$ is $K$-close to $y$ where $K \leq n$.  Thus, $x_1$ is $(K-1)$-close, $x_2$ is $(K-2)$-close, and so on until $x_K$ is $0$-close.  That $x_K$ is a closest lattice point to $y$ follows from Lemma~\ref{lem:rngzeroclosestpoint}.
\end{proof}

We now provide proof of Lemmas~\ref{lem:vqcontramin} and~\ref{lem:vqcontramax} that were used in the proof of Lemma~\ref{lem:rngdecreases}.  We first require the following two simple Lemmas.

\begin{lemma}\label{lem:simplevminh}
Let $v \in \ints^{n+1}$ and $h \in \{0,1\}^{n+1}$ be such that $h_i = 0$ whenever $v_i = \min(v)$.  If $S \subseteq \{ i \mid v_i = \min(v) \}$ then $h_i - h_j \geq v_i - v_j$ for all $i \in S$ and $j = 1,\dots,n+1$.
\end{lemma}
\begin{proof}
For $i \in S$ we have $v_i = \min(v)$ and $h_i = 0$ and so the lemma is true if we show that $-h_j \geq \min(v) - v_j$ for all $j=1,\dots,n+1$.  If $v_j = \min(v)$ then $h_j = 0$ so that $-h_j = 0 = \min(v) - v_j$.  Otherwise, if $v_j \neq \min(v)$ then 
\[
-h_j \geq -1 \geq \min(v) - v_j
\] 
because $h_j \in \{0,1\}$ and $v_j$ and $\min(v)$ are integers with $v_j > \min(v)$.
\end{proof}

\begin{lemma}\label{lem:simplevmaxh}
Let $v \in \ints^{n+1}$ and $h \in \{0,1\}^{n+1}$ be such that $h_i = 1$ whenever $v_i = \max(v)$.  If $S \subseteq \{ i \mid v_i = \max(v) \}$ then $h_j - h_i \geq v_j - v_i$ for all $i \in S$ and $j = 1,\dots,n+1$.
\end{lemma}
\begin{proof}
For $i \in S$ we have $v_i = \max(v)$ and $h_i = 1$ and so the lemma is true if we show that $h_j-1 \geq v_j - \max(v)$ for all $j=1,\dots,n+1$.  If $v_j = \max(v)$ then $h_j = 1$ and $h_j-1 = 0 = v_j - \max(v)$.  Otherwise, if $v_j \neq \max(v)$ then 
\[
h_j-1 \geq -1 \geq v_j - \max(v)
\] 
because $h_j \in \{0,1\}$ and $v_j$ and $\max(v)$ are integers with $v_j < \max(v)$.
\end{proof}
